---
title: "Bike Sharing Analysis"
author: "Irvin Romero"
date: "9/25/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Data Preparation

Load in necessary libraries & set working directory

```{r cars, results = FALSE, warning=FALSE, message=FALSE}
rm(list = ls())
library(data.table)
library(tidyverse)  
library(lubridate)  
library(ggplot2)  
library(ggthemes)
library(plotly)

setwd("~/Documents/bikeSharing")


```


Since our data is stored across multiple csv files we need to extract and combine all together

```{r}
file_names <- dir("data", pattern = "csv", full.names = TRUE) 
all_trips <- rbindlist(lapply(file_names, fread))
```

Calculating ride length and creating new time range variables

```{r}
all_trips$ride_time_secs <- difftime(all_trips$ended_at, all_trips$started_at, units = "secs")
all_trips$wday <- wday(all_trips$started_at)
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$month <- format(as.Date(all_trips$date), "%m")

colnames(all_trips)[13] <- "rider_type"

```
We can remove trips performed by HQ and test rides
```{r}
all_trips <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_time_secs < 0),]


```



Our data contains 3.5 million rides between 4/2020 and 4/2021 totaling to over 5.8 billion seconds (97 mil mins.) of ride time

```{r}
all_trips %>% 
  summarise(number_of_rides = n(), total_ride_sec = sum(ride_time_secs),  start_date =min(date), end_date = max(date)) 
```

### How do Cyclistic Members and Casual Riders use Cyclistic Bikes differently?


Comparing the Total of Cyclistic members and Casual riders
```{r }
table(all_trips$rider_type)
```

Casual users on average ride for 45 mins (2698 secs) 
Members on average ride for 16 mins (966 secs)
Median ride time is twice as much for casual riders vs member riders

```{r}
all_trips %>%
  group_by(rider_type) %>% 
  summarise(Mean_ride_time_secs = mean(ride_time_secs),
            Median_ride_time_secs = median(ride_time_secs)) %>% 
  data.frame()
```
### Member riders are more likely to ride Cyclistic bikes daily compared to casual riders who are more likely to ride on the weekends

```{r, message=FALSE}
x <- all_trips %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  
  group_by(rider_type, weekday) %>%  
  summarise(Number_of_rides = n()							 
            ,Average_duration = mean(ride_time_secs)
            ,Total_duration = sum(ride_time_secs)) 
  
countPlot <- ggplot(x) +
  geom_col(aes(x = weekday, y = Number_of_rides, fill = rider_type), position = "dodge") + 
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  scale_fill_manual(values = c("dodgerblue2", "green4")) + ggtitle("Total Rides by Day")+
  xlab("Weekday") + ylab("Number of Rides") + theme_classic()

ggplotly(countPlot)
```

## Daily average ride time shows casual riders ride 3 times longer than member riders

```{r}
dailyPlot <- ggplot(x)  +
  geom_col(aes(x = weekday, y = Average_duration, fill = rider_type),position = "dodge")+
  scale_fill_manual(values = c("dodgerblue2", "green4")) + ggtitle("Average Duration by Day")+
  xlab("Weekday") + ylab("Avg Duration") + theme_classic()

ggplotly(dailyPlot)
```

Casual and Member riders both ride docked bikes over other options

```{r}
countTypesPlot <- all_trips %>% 
  group_by(rider_type, rideable_type) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_time_secs)) %>% 
  arrange(rider_type, rideable_type)  %>% 
  ggplot(aes(x = rideable_type, y = number_of_rides, fill = rider_type)) +
  geom_col(position = "dodge")+
  scale_fill_manual(values = c("dodgerblue2", "green4")) + ggtitle("Count of rides by Type of Transport")+
  xlab("Transportation Type") + ylab("Count of Rides") + theme_classic()

ggplotly(countTypesPlot)
```

Casual riders on average ride docked bikes for longer periods of time compared to other bike options

```{r}
durationPlot <- all_trips %>% 
  group_by(rider_type, rideable_type) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_time_secs)) %>% 
  arrange(rider_type, rideable_type)  %>% 
  ggplot(aes(x = rideable_type, y = average_duration, fill = rider_type)) +
  geom_col(position = "dodge")+
  scale_fill_manual(values = c("dodgerblue2", "green4")) + ggtitle("Avg Duration by Type of Transport")+
  xlab("Transportation Type") + ylab("Avg duration (secs)") + theme_classic()

ggplotly(durationPlot)
```


```{r}
dateRides <- all_trips %>% 
  group_by( rider_type, month) %>% 
  summarise( total_ride_dura = sum(ride_time_secs)) 
x <- ggplot(dateRides) + 
  geom_col(aes( x = month, y = as.numeric(total_ride_dura), fill= rider_type), position = "dodge")+
  scale_fill_manual(values = c("dodgerblue2", "green4")) + ggtitle("Avg Duration by Type of Transport")+
  xlab("Transportation Type") + ylab("Avg duration (secs)") + theme_classic()

ggplotly(x)
```



Going forward I would take a deeper look into what locations had the highest amount of casual riders to get a better understanding of places we could target for our marketing campaigns 



